\opage{
\otitle{9.3 The energy of a single Slater determinant}

\otext

We will first express the Slater determinant ($\Phi$) in terms of the anti-symmetrizer operator $\hat{A}$ and the simple
product of orbitals $\phi_i$ ($\Pi$):

\aeqn{9.5}{\Phi = \hat{A}\left(\phi_1(1)\phi_2(2)...\phi_n(n)\right) = \hat{A}\Pi}

The antisymmetrizer operator is defined as:

\aeqn{9.6}{\hat{A} = \frac{1}{\sqrt{n!}}\sum_{p=0}^{n-1}\left(-1\right)^p\hat{P} = \frac{1}{\sqrt{n!}}\left(\hat{1}
-\sum_{i,j}\hat{P}_{i,j} + \sum_{i,j,k}\hat{P}_{i,j,k} - ...
\right)}

where $\hat{1}$ is the identity operator, $\hat{P}_{i,j}$ generates all possible permutations of two electron coordinates, 
$\hat{P}_{i,j,k}$ all possible permutations of three electron coordinates, etc. The following results for $\hat{A}$ can be shown
to hold:

\aeqn{9.10}{\hat{A}\hat{H} = \hat{H}\hat{A}\textnormal{ (commutes with the Hamiltonian)}}
\aeqn{9.11}{\hat{A}\hat{A} = \sqrt{n!}\hat{A}}

The total Hamiltonian can be broken down as follows:

\aeqn{9.12}{\hat{H} = \hat{T}_e + \hat{V}_{ne} + \hat{V}_{ee} + \hat{V}_{nn}}

where $\hat{T}_e$ is the electron kinetic energy, $\hat{V}_{ne}$ is the nuclear-electron attraction, $\hat{V}_{ee}$ is the
electron-electron repulsion, and $\hat{V}_{nn}$ is the nuclear-nuclear repulsion (constant for given molecular geometry).

}

\opage{

\otext

The component operators above are defined as ($N$ is the number of nuclei in the molecule):

\vspace*{-0.5cm}

\deqn{9.13}{\hat{T}_e = -\sum_{i=1}^n\frac{1}{2}\nabla_i^2}
{\hat{V}_{ne} = -\sum_{i=1}^{n}\sum_{j=1}^{N}\frac{Z_j}{\left|R_j-r_i\right|}}
{\hat{V}_{ee} = \sum_{i < j}^{n}\frac{1}{\left|r_i - r_j\right|}}
{V_{nn} = \sum_{i < j}^{N}\frac{Z_iZ_j}{\left|R_i - R_j\right|}}

Since the nuclear - nuclear repulsion term $V_{nn}$ does not depend on the electron coordinates, it can be added to the total energy after the calculation.
The remaining electronic part ($\hat{H}_e$) can be written in terms of the one and two electron operators:

\vspace*{-0.25cm}

\beqn{9.14}{\hat{H}_e = \sum_{i=1}^n\hat{h}_i + \sum_{i < j}^n\hat{g}_{ij} + V_{nn}}
{\hat{h}_i = -\frac{1}{2}\nabla^2_i - \sum_{j=1}^{N}\frac{Z_j}{\left|R_j - r_i\right|}\textnormal{ and }\hat{g}_{ij} = \frac{1}{\left|r_i - r_j\right|}}

}

\opage{

\otext

The energy of a Slater determinant can be writen in terms of the permutation operator $P$ as follows:

\aeqn{9.15}{E = \left<\Phi\left|\hat{H}\right|\Phi\right> = \left<\hat{A}\Pi\left|\hat{H}\right|\hat{A}\Pi\right> = \sqrt{n!}\left<\Pi\left|\hat{H}\right|\hat{A}\Pi\right> = \sum_p\left(-1\right)^p\left<\Pi\left|\hat{H}\right|\hat{P}\Pi\right>}

For the one-electron $\hat{h}_1$ ($i = 1$; other indices evaluated the same way) and the above becomes:

\beqn{9.16}{\left<\Pi\left|\hat{h}_1\right|\Phi\right> = \left<\phi_1(1)\phi_1(2)...\phi_n(n)\left|\hat{h}_1\right|\phi_1(1)\phi_2(2)...\phi_n(n)\right>}
{\left<\phi_1(1)\left|\hat{h}_1\right|\phi_1(1)\right>\left<\phi_2(2)|\phi_2(2)\right>\left<\phi_3(3)|\phi_3(3)\right>...\left<\phi_n(n)|\phi_n(n)\right>}

Note that only the identity operator in the permutation operator of Eq. (\ref{eq9.6}) can give a non-zero contribution. All actual permutation operators
involve change in the electron indices and therefore they include terms such as $\left<\phi_2(2)|\phi_1(2)\right>$ etc., 
which are zero based on orthogonality. Consider, for example, the following two-electron permutation operator:

\beqn{9.17}{\left<\Pi\left|\hat{h}_1\right|\hat{P}_{12}\Pi\right> = \left<\phi_1(1)\phi_2(2)...\phi_n(n)\left|\hat{h}_1\right|\phi_2(1)\phi_1(2)...\phi_n(n)\right>}
{=\left<\phi_1(1)\left|\hat{h}_1\right|\phi_2(1)\right>\left<\phi_2(2)|\phi_1(2)\right>...\left<\phi_n(n)|\phi_n(n)\right>}

}

\opage{

\otext

For the two-electron operator ($g_{ij}$), both the identity and $\hat{P}_{ij}$ operators can give non-zero contributions. However, 
three and higher permutation operators create at least one overlap integral between two different orbitals (orthogonality) and
hence these must be zero. The identity operator yields the following contribution (\textit{Coulomb integral}):

\vspace*{-0.25cm}

\ceqn{9.18}{\left<\Pi\left|\hat{g}_{12}\right|\Pi\right> = \left<\phi_1(1)\phi_2(2)...\phi_n(n)\left|\hat{g}_{12}\right|\phi_1(1)\phi_2(2)...\phi_n(n)\right>}
{= \left<\phi_1(1)\phi_2(2)\left|\hat{g}_{12}\right|\phi_1(1)\phi_2(2)\right>\left<\phi_3(3)|\phi_3(3)\right>...\left<\phi_n(n)|\phi_n(n)\right>}
{= \left<\phi_1(1)\phi_2(2)\left|\hat{g}_{12}\right|\phi_1(1)\phi_2(2)\right> = J_{12}}

This represents the classical Coulomb interaction between the two electrons on these orbitals. The two-electron permutation operator yields (\textit{Exchange integral}):

\ceqn{9.19}{\left<\Pi\left|\hat{g}_{12}\right|\hat{P}_{12}\Pi\right> = \left<\phi_1(1)\phi_2(2)...\phi_n(n)\left|\hat{g}_{12}\right|\phi_2(1)\phi_1(2)...\phi_n(n)\right>}
{= \left<\phi_1(1)\phi_2(2)\left|\hat{g}_{12}\right|\phi_2(1)\phi_1(2)\right>\left<\phi_3(3)|\phi_3(3)\right>...\left<\phi_n(n)|\phi_n(n)\right>}
{= \left<\phi_1(1)\phi_2(2)\left|\hat{g}_{12}\right|\phi_2(1)\phi_1(2)\right> = K_{12}}

The above quantities can be arranged in matrix form (vector $h_i$ \& matrices $J$ and $K$) according to electron indices. The total energy corresponding
to a given Slater determinant can then be obtained from:

\vspace*{-0.25cm}

\beqn{9.20}{E = \sum_{i=1}^nh_i + \sum_{i=1}^n\sum_{j>i}^n\left(J_{ij} - K_{ij}\right) + V_{nn}}
{= \sum_{i=1}^nh_i + \frac{1}{2}\sum_{i=1}^n\sum_{j=1}^n\left(J_{ij} - K_{ij}\right) + V_{nn}}

}

\opage{

\otext

Note that the minus sign in front of $K_{ij}$ above originates from the $(-1)^p$ term in Eq. (\ref{eq9.6}). This can also be 
expressed in terms of Coulomb and Exchange operators:

\aeqn{9.21}{E = \sum_{i=1}^n\left<\phi_i\left|\hat{h}_i\right|\phi_i\right> + \frac{1}{2}\sum_{i,j}^n\left(\left<\phi_j\left|\hat{J}_i\right|\phi_j\right> - \left<\phi_j\left|\hat{K}_i\right|\phi_j\right>\right) + V_{nn}}

The operators $\hat{J}_i$ and $\hat{K}_i$ are defined as:

\beqn{9.22}{\hat{J}_i|\phi_j(2)> = \left<\phi_i(1)\left|\hat{g}_{12}\right|\phi_i(1)\right>|\phi_j(2)>}
{\hat{K}_i|\phi_j(2)> = \left<\phi_i(1)\left|\hat{g}_{12}\right|\phi_j(1)\right>|\phi_i(2)>}

Given the molecular orbitals, this expression yields the electronic energy for the atom/molecule.

}
