\documentclass{report}

\begin{document}
\setlength{\parindent}{0pt}
\textbf{SYLLABUS}\\

\begin{tabular}{ll}
Course:   & CHEM 352 (Quantum Chemistry, 4 units)\\
Prerequisites: & CHEM351; PHYS 226 or 220B; MATH 150B or 255B\\
Lecturer: & Dr. Jussi Eloranta\\
E-mail:	  & Jussi.Eloranta@csun.edu\\
Office:	  & Eucalyptus Hall 2025\\
Office hours: & Mondays 2:00 -- 3:00 pm.\\
Lectures: & Mon, Wed 8:00 am -- 9:40 am. 4 hours / week of lectures\\
          & in EH2028\\
Exams: & One midterm (weight 50 \%) and a final (weight 50 \%)\\
Content: & Quantum mechanics, quantum chemistry, optical spectroscopy,\\
         & magnetic resonance spectroscopy\\
Optional material: & Physical Chemistry by P. W. Atkins and J. de Paula,\\
                   & Physical Chemistry by Silbey, Alberty and Bawendi\\
Web page: & \verb+http://www.csun.edu/~jeloranta/CHEM352/+\\
\end{tabular}

\vspace{0.5cm}

\textbf{1. Table of contents}\\

\begin{tabular}{l@{\extracolsep{3cm}}l}
1. & Introduction to quantum mechanics\\
2. & Quantum mechanics of atoms\\
3. & Quantum mechanics of molecules\\
4. & Group theory\\
5. & Optical spectroscopy\\
6. & Electronic spectroscopy (optical spectroscopy continued)\\
7. & Magnetic resonance spectroscopy\\
\end{tabular}

\vspace*{0.4cm}

\textbf{2. Tentative schedule (spring 2009)}\\

\begin{tabular}{l@{\extracolsep{3cm}}l}
\underline{Chapters} & \underline{Examination}\\
1 -- 3 & Midterm (target March 17)\\
4 -- 6 & Final examination (May 16)\\
\end{tabular}

\vspace*{0.4cm}

\textbf{3. Homework}\\

Additional examples/homework for each chapter can be downloaded from the course web page.

\vspace*{0.4cm}

\textbf{4. Examinations}\\

Additional material is allowed in the examinations (including lecture notes, textbooks, programmable calculators, 
etc.). A \textit{tentative} grading scale is as follows:

\begin{center}
\begin{tabular}{l@{\extracolsep{3cm}}l}
\underline{Grade}			&	\underline{Examination score}\\
A			&	90 -- 100 points\\
B			&	75 -- 90 points\\
C			&	65 -- 75 points\\
D			&	50 -- 65 points\\
F			&	$<$ 50 points\\
\end{tabular}
\end{center}

The overall grade is taken as a weighted average of the examination scores.\\

\vspace{0.5cm}

\textbf{5. Suggested reference material}\\

The following reference material will be helpful during the course:

\begin{enumerate}
\item Physics Handbook for Science and Engineering, C. Nordling and J. \"Osterman, Studentlitteratur (2004).
\item Mathematics for Physical Chemistry (3rd ed.), R. G. Mortimer, Academic Press (2005).
\end{enumerate}

\vspace{0.5cm}

\textbf{6. Practical hints}

\begin{enumerate}
\item Read the corresponding textbook section and the notes before the lectures. The notes are available at
the course web page. Ask questions!
\item The best way to learn physical chemistry is through exercises. This is the reason for the homework 
being mandatory.
\item Always try to understand the whole concept first and the work out the details.
\item Try to understand the material instead of just memorizing it. The latter approach will not work in 
physical chemistry.
\end{enumerate}

\vspace*{0.2cm}

\textbf{7. Academic dishonesty}\\

By enrolling in this class, you agree to abide by all California State University, Northridge policies of academic honesty and integrity. Students violating these standards will receive a zero for the work in question and will have their case referred to the Student Affairs Office for appropriate disciplinary action. See pages 586-589 of the 2008-2010 California State University, Northridge catalog for details of the University policies.


\end{document}
