\noindent
\textbf{Thermodynamics:
\ifthenelse{\equal{\solutions}{true}}{Examples}{Homework} for chapter 4.}\\

\begin{enumerate}

\item One mole of nitrogen gas is allowed to expand from 0.5 to 10 L (reversible and isothermal process at 300 K). Calculate the change in molar entropy using (a) the ideal
gas law and (b) the van der Waals equation with $a = 1.408$ atm L$^2$ mol$^{-2}$, $b = 0.03913$ L mol$^{-1}$.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-1}}{}

\item Derive the relation for $\bar{C}_P - \bar{C}_V$ for a gas that follows the van der Waals equation.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-2}}{}

\item Estimate the change in molar entropy of liquid benzene at 25 $^\circ$C and atmospheric pressure when the pressure is raised to 1000 bar? The coefficient of thermal expansion $\alpha$ is 1.237 $\times$ 10$^{-3}$ K$^{-1}$, the density is 0.879 g cm$^{-3}$, and the molar mass is 78.11 g mol$^{-1}$.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-3}}{}

\item Derive the expression for $\left(\partial U / \partial V\right)_T$ (the internal pressure) for a gas following the virial equation with $Z = 1 + B / V$. Consider only $PV$-work and a reversible process. Note that $B = B(T)$ (i.e. it depends on temperature).

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-4}}{}

\item Consider the process of freezing water at $-10$ $^\circ$C for which the enthalpy change is $-5619$ J mol$^{-1}$ and entropy change is $-20.54$ J K$^{-1}$ mol$^{-1}$. What is the Gibbs energy of freezing water at constant temperature of $-10$ $^\circ$C?

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-5}}{}

\item (a) Integrate the Gibbs-Helmholtz to obtain an expression for $\Delta G_2$ at temperature $T_2$ in terms of $\Delta G_1$ at $T_1$, assuming that $\Delta H$ is independent of temperature. (b) Obtain an expression for $\Delta G_2$ using a more accurate approximation that $\Delta H = \Delta H_1 + (T - T_1)\Delta C_P$, where $T_1$ is an arbitrary reference temperature. Assume that $\Delta C_P$ is temperature independent. Hint: Integrate one side with respect to variable $\Delta G / T$ and the other with respect to $T$.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-6}}{}

\item An ideal gas is allowed to expand reversibly and isothermally (25 $^\circ$C) from a pressure of 1 bar to a pressure of 0.1 bar. (a) What is the change in molar Gibbs energy? (b) What would be the change in molar Gibbs energy if the process occurred irreversibly?

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-7}}{}

\item Helium is compressed isothermally and reversibly at 100 $^\circ$C from a pressure of 2 to 10 bar. Calculate (a) $q$ per mole, (b) $w$ per mole, (c) $\Delta\bar{G}$, (d) $\Delta\bar{A}$, (e) $\Delta\bar{H}$, (f) $\Delta\bar{U}$, (g) $\Delta\bar{S}$, assuming that helium is an ideal gas.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-8}}{}

\item Toluene (molecular weight 92.13 g mol$^{-1}$) is vaporized at constant external pressure and its boiling point, 111 $^\circ$C (constant temperature). The heat of vaporization at this temperature is 361.9 J g$^{-1}$. For the vaporization of toluene, calculate (a) $w$ per mole, (b) $q$ per mole, (c) $\Delta\bar{H}$, (d) $\Delta\bar{U}$, (e) $\Delta\bar{G}$, and (f) $\Delta\bar{S}$. In part (a) assume that the volume of the liquid is negligible and that toluene vapor behaves according to the ideal gas law (this
assumption is not needed elsewhere). Note also that evaporation at the boiling point is a reversible process.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-9}}{}

\item Calculate $\Delta_{mix} G$ and $\Delta_{mix} S$ for the formation of a quantity of air containing 1 mol of gas by mixing nitrogen and oxygen at 298.15 K. Air may be taken to be 80\% nitrogen and 20\% oxygen (in mol \%). Assume ideal gas behavior.

\ifthenelse{\equal{\solutions}{true}}{\input{ch4-solution-10}}{}

\end{enumerate}
