\noindent
\textbf{CHEM 352:
\ifthenelse{\equal{\solutions}{true}}{Examples}{Homework} for chapter 2.}\\

\begin{enumerate}

% Problem 1

\item The quantum mechanical state of a hydrogen atom is described by the following superposition:
$$\psi = \frac{1}{\sqrt{14}}\left(2\psi_{1,0,0} - 3\psi_{2,0,0} - \psi_{3,2,2}\right)$$

where $\psi_{n,l,m}$ are eigenfunctions of the Hamiltonian. The subscripts refer to quantum numbers $n,l,m$.

\begin{enumerate}
\item What is the probability of finding the hydrogen atom in states $(n = 1, l = 0, m = 0), (n = 2, l = 0, m = 0), (n = 3, l = 2, m = 2)$ or in some other state?
\item What are the expectation values for energy, $\vec{\hat{L}}^2$ and $\hat{L}_z$?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-1}}{}

% Problem 2

\item Show that operators $\hat{L}_z$ and $\vec{\hat{L}}^2 = \hat{L}_x^2 + \hat{L}_y^2 + \hat{L}_z^2$ commute with the hydrogen atom Hamiltonian operator:
$$\left[-\frac{\hbar^2}{2m}\Delta + \hat{V}, \hat{L}_z\right] = \left[-\frac{\hbar^2}{2m}\Delta + \hat{V}, \vec{\hat{L}}^2\right] = 0$$

where $\hat{V}$ is the operator corresponding to electron - nuclear Coulomb interaction. Use spherical coordinates and remember that operators commute, for example, if they depend on different variables. What is the significance of this result?

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-2}}{}

% Problem 3

\item Demonstrate that the Cartesian hydrogen like $p_x$ and $p_y$ orbitals are not eigenfunctions of $\hat{L}_z$ but their linear combinations $p_x \pm ip_y$ are.

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-3}}{}

% Problem 4

\item 

\begin{enumerate}
\item Consider a hydrogenlike atom with on electron on $2s$ orbital. What is the most probable distance from the nucleus? Use the radial wavefunction in your calculation.
\item Show that the following hydrogenlike atom orbital pairs are orthogonal: ($1s,2s$) and ($2p_x, 2p_y$).
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-4}}{}

% Problem 5

\item 

\begin{enumerate}
\item Write the electron configuration for V$^{2+}$ ion. What quantum numbers for the total electron spin are possible in this configuration?
\item If two electrons reside on two different orbitals, what are the possible values for total spin and the multiplicity? What values are possible for three electrons on different orbitals?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-5}}{}

% Problem 6

\item 

\begin{enumerate}
\item What information do the following term symbols provide about a given atom: $^1D_2$ and $^3F_4$?
\item Consider the emission spectrum of potassium atom, which exhibits lines at $\lambda_1 = 766.70$ nm and $\lambda_2 = 770.11$ nm. What is the value of the spin-orbit coupling constant? The emission lines originate from the $^2P$ excited state (spin-orbit split).
\item Which of the following atomic transitions are (dipole) allowed: $5d\rightarrow 2s, 5p \rightarrow 3s, 5p \rightarrow 3f$?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-6}}{}

% Problem 7

\item

\begin{enumerate}
\item Write all the term symbols that can be obtained from the following electron configurations: $2s^12p^1$, $2p^13d^1$ and Ar$4s^23d^{10}4p^5$ (Br atom).
\item Write the term symbols for carbon atom (He$2d^22p^2$), which has two equivalent $p$-electrons. Hint: tabulate all possible $m_{l_1}, m_{l_2}, m_{s_1}$ and $m_{s_2}$ values and calculate the total $L$ and $S$ values. Remember to exculde the Pauli forbidden states.
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch2-solution-7}}{}

\end{enumerate}
