\noindent
\textbf{CHEM 352:
\ifthenelse{\equal{\solutions}{true}}{Examples}{Homework} for chapter 5.}\\

\noindent
1. 
\begin{itemize}
\item[a)] The molar absorption coefficient of a substance dissolved in hexane is $\epsilon = 855$ L mol$^{-1}$ cm$^{-1}$ at $\lambda = 270$ nm. Calculate the intensity reduction in percentage when light passes through a 2.5 mm thick film of 3.25 mmol/L solution.

\item[b)] Consider a 10 mmol/L solution of benzene in a non-absorbing solvent. The solution was placed in a 2.0 mm thick cuvette and the transmission of 256 nm light through the sample was observed to be 48 \%. What is the molar absorption coefficient of benzene at 256 nm? What would be the transmittance when using a 4.0 mm thick cuvette at the same wavelength?

\end{itemize}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-1}}{}

\noindent
2. Compare the ratio $A / B$ between the Einstein spotaneous and stimulated emission coefficients for the following wavelengths: a) $\lambda$ = 70.8 pm (X-ray), b) $\lambda$ = 500 nm (visible light), c) $\tilde{\nu}$ = 3000 cm$^{-1}$ (IR), d) $\lambda$ = 3 cm (microwaves), e) $\nu$ = 500 MHz (radiowaves). What does this tell you about the significance of the spontaneous emission at different energies?\\

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-2}}{}

\noindent
3. 

\begin{itemize}

\item[a)] Calculate the relative Doppler broadening for gaseous ICl molecules at 25 $^\circ$C. What are the linewidths $\delta\nu_{rot}$ (kHz) and $\delta\nu_{vib}$ (cm$^{-1}$) when the rotational constant $B = 0.1142$ cm$^{-1}$ and the vibrational frequency is $\nu = 384$ cm$^{-1}$.

\item[b)] If the excited state has a lifetime of 100 ps, what is the lifetime broadening caused by this?

\end{itemize}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-3}}{}

\noindent
4. The rotational spectrum of $^{127}$I$^{35}$Cl shows lines with 0.2284 cm$^{-1}$ spacings. What is the bond length of this molecule?\\

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-4}}{}

\noindent
5. Consider NH$_3$ molecule (non-planar geometry; symmetric top).

\begin{itemize}
\item[a)] What are the positions of the four first Stokes and anti-Stokes rotational Raman lines when the excitation laser wavelength is 336.732 nm and the rotational constant $B$ = 9.977 cm$^{-1}$ (note that you do not need the rotational constant $A$ in this calculation since the selection rule include $\Delta K = 0$).

\item[b)] Demonstrate that the above rotational constant is consistent with N-H bondlength of 101.2 pm and bond angle 106.7$^\circ$. 

\end{itemize}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-5}}{}



