\noindent
\textbf{Thermodynamics:
\ifthenelse{\equal{\solutions}{true}}{Examples}{Homework} for chapter 5.}\\

\begin{enumerate}

\item For the reaction N$_2(g) + 3$H$_2(g) = 2$NH$_3(g)$, $K = 1.60 \times 10^{-4}$ at 400 $^\circ$C. Assume ideal gas behavior for the gases. Calculate:

\begin{enumerate}
\item $\Delta_rG^\circ$
\item $\Delta_rG$ when the partial pressures of N$_2$ and H$_2$ are maintained at 10.0 and 30.0 bar, respectively, and NH$_3$ is removed at a partial pressure of 3 bar.
\item Is the reaction spontaneous under the latter conditions?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch5-solution-1}}{}

\item At 55 $^\circ$C and 1 bar the average molar mass of partially dissociated N$_2$O$_4$ is 61.2 g mol$^{-1}$. The molecular mass of pure N$_2$O$_4$ is 92.01 g mol$^{-1}$. Assume ideal gas behavior.

\begin{enumerate}
\item Calculate the extent of reaction ($\xi$).
\item $K$ for the reaction N$_2$O$_4(g) = 2$NO$_2(g)$.
\item Calculate $\xi$ at 55 $^\circ$C if the total pressure is reduced to 0.1 bar.
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-2}}{}

\item Express $K$ for the reaction: CO$(g) + 3$H$_2(g) =$ CH$_4(g) + $H$_2$O$(g)$ in terms of the equilibrium extent of the reaction $\xi_{eq}$ when one mole of CO is mixed with one mole of hydrogen. Assume ideal gas behavior.

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-3}}{}

\item At 1273 K and at a total pressure of 30.4 bar the equilibrium in the reaction CO$_2(g) + $C$(s) = 2$CO$(g)$ is such that 17 mol \% of the gas is CO$_2$. Assume ideal behavior for gases and activities for solids can be taken to be one.

\begin{enumerate}
\item What percentage would be CO$_2$ if the total pressure were 20.3 bar?
\item What would be the effect on the equilibrium of adding N$_2$ to the reaction mixture in a closed vessel until the partial pressure of N$_2$ is 10 bar?
\item At what pressure of the reactants will 25\% of the gas be CO$_2$?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-4}}{}

\item The following reaction is nonspontaneous at room temperature and endothermic:

$$3\textnormal{C}(\textnormal{graphite}) + 2\textnormal{H}_2\textnormal{O}(g) = \textnormal{CH}_4(g) + 2\textnormal{CO}(g)$$

As the temperature is raised, the equilibrium constant will become equal to unity at some point. Estimate this point using the data given below (in units of kJ mol$^{-1}$). Finally, use the van't Hoff equation with assumption that $\Delta_rH^\circ$ is independent of temperature to get a better estimate for this temperature.

\begin{tabular}{lllll}
$T$ (K) & $\Delta_fG^\circ($CH$_4(g))$ & $\Delta_fG^\circ($CO$(g))$ & $\Delta_fG^\circ($H$_2$O$(g))$ & $\Delta_rG^\circ($C$(s))$\\
0    & -66.911 & -113.805 & -238.921 & 0.000\\
298  & -50.768 & -137.163 & -228.582 & 0.000\\
500  & -32.741 & -155.414 & -219.051 & 0.000\\
1000 & 19.492  & -200.275 & -192.590 & 0.000\\
2000 & 130.802 & -286.034 & -135.528 & 0.000\\
3000 & 242.332 & -367.816 & -77.163  & 0.000\\
\end{tabular}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-5}}{}

\item Starting with the fundamental equation for $G$ in the form: $dG = -SdT + VdP + \Delta_rGd\xi$, derive equations for calculating $\Delta_rS$, $\Delta_rV$ and $\Delta_rH$ in terms of $\Delta_rG(P,T)$ for a chemical reaction. Hint: Use the Maxwell relations.

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-6}}{}

\item The decomposition of silver oxide is represented by 2Ag$_2$O$(s) = 4$Ag$(s) + $O$_2(g)$. Use the NIST Chemistry Webbook database (for $\Delta_fH^\circ$ and $\bar{S}^\circ$) and assume $\Delta_rC_P^\circ = 0$ to calculate the temperature at which the equilibrium pressure of O$_2$ is 0.2 bar. This temperature is of interest because Ag$_2$O will decompose to yield Ag at temperatures above this value if it is in contact with air. Assume that O$_2$ follows the ideal gas law and that the activities of solid reagents are one.

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-7}}{}

\item The reaction 2NOCl$(g) = 2$NO$(g) + $Cl$_2(g)$ comes to equilibrium at 1 bar total pressure and 227 $^\circ$C when the partial pressure of the nitrosyl chloride (NOCl) is 0.64 bar. Only NOCl was present initially. Assume ideal gas behavior.

\begin{enumerate}
\item Calculate $\Delta_rG^\circ$ for this reaction.
\item At what total pressure will the partial pressure of Cl$_2$ be 0.1 bar?
\end{enumerate}

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-8}}{}

\item For a chemical reaction, $\ln(K) = a + b / T + c / T^2$. Derive the corresponding expressions to calculate $\Delta_rG^\circ$, $\Delta_rH^\circ$, $\Delta_rS^\circ$, and $\Delta_rC_P^\circ$.

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-9}}{}

\item The equilibrium constant for the association of benzoic acid to a dimer in dilute benzene solutions is as follows at 43.9 $^\circ$C: 2C$_6$H$_5$COOH$ = ($C$_6$H$_5$COOH$)_2$ with $K_c = 2.7 \times 10^2$. Note that molar concentrations were used in expressing the equilibrium constant. Calculate $\Delta_rG^\circ$ and state its meaning. Hint: In dilute solutions activities can be replaced by concentrations.

\ifthenelse{\equal{\solutions}{true}}{\input{ch5-solution-10}}{}

\end{enumerate}
