\noindent
\textbf{Thermodynamics:
\ifthenelse{\equal{\solutions}{true}}{Examples}{Homework} for chapter 6.}\\

\begin{enumerate}

\item The boiling point of hexane at 1 atm is 68.7 $^\circ$C. What is the boiling point at 1 bar? The vapor pressure of hexane at 49.6 $^\circ$C is 53.32 kPa. Assume that the vapor phase obeys the ideal gas law and that the molar volume of the liquid is negligible compared to the molar volume of gas.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-1}}{}

\item Liquid mercury has a density of 13.690 g cm$^{-3}$ and solid mercury 14.193 g cm$^{-3}$, both measured at the melting point ($-38.87$ $^\circ$C) and 1 bar pressure. The heat of fusion is 9.75 J g$^{-1}$. Calculate the melting points of mercury under a pressure of (a) 10 bar and (b) 3540 bar. The experimentally observed melting point under 3540 bar is $-19.9$ $^\circ$C.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-2}}{}

\item From the $\Delta_rG^\circ$ (3110 J mol$^{-1}$ for the gas) of Br$_2$(g) at 25 $^\circ$C, calculate the vapor pressure of Br$_2$(l). The pure liquid at 1 bar and 25 $^\circ$C is taken as the standard state (e.g., $\Delta_fG^\circ = 0$ for the liquid). Assume that Br$_2$(g) follows the ideal gas law.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-3}}{}

\item A binary mixture of A and B is in equilibrium with its vapor at constant temperature and pressure. Prove that $\mu_A(g) = \mu_{A}(l)$ and $\mu_B(g) = \mu_B(l)$ by starting with $G = G(g) + G(l)$ and the fact that $dG = 0$ when infinitesimal amounts of A and B are simultaneously transferred from the liquid to the vapor.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-4}}{}

\item One mole of benzene (component 1) is mixed with two moles of toluene (component 2). At 20 $^\circ$C, the vapor pressures of benzene and toluene are 51.3 and 18.5 kPa, respectively. (a) As the pressure is reduced, at what pressure will boiling begin? (b) What will be the composition of the first bubble of vapor? Assume that the mixture follows Raoult's law and that the gas phase is ideal.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-5}}{}

\item Calculate the solubility of naphthalene at 25 $^\circ$C in any solvent in which it forms an ideal solution. The melting point of naphthalene is 80 $^\circ$C, and the enthalpy of fusion is 19.29 kJ mol$^{-1}$. The measured solubility of naphthalene in benzene is $x_1 = 0.296$. Assume that this forms an ideal solution.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-6}}{}

\item The addition of a nonvolatile solute to a solvent increases the boiling point above that of the pure solvent. What is the elevation of the boiling point when 0.1 mol of nonvolatile solute is added to 1 kg of water? The enthalpy of vaporization of water at the boiling point is 40.6 kJ mol$^{-1}$. Assume that this forms an ideal solution.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-7}}{}

\item For a solution of water (component 1) and ethanol (component 2) at 20 $^\circ$C that has 0.2 mole fraction of ethanol, the partial molar volume of water is 17.9 cm$^{3}$mol$^{-1}$ and the partial molar volume of ethanol is 55.0 cm$^3$mol$^{-1}$. What volumes of pure ethanol and water are required to make exactly 1 L of this solution? At 20 $^\circ$C the density of ethanol is 0.789 g cm$^{-3}$ and the density of water is 0.998 g cm$^{-3}$. The molecular weights of water and ethanol are 18.016 and 46.07 g mol$^{-1}$, respectively.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-8}}{}

\item Calculate $\Delta_rG^\circ$ for H$_2$O(g, 25 $^\circ$C) = H$_2$O(l, 25 $^\circ$C). The equilibrium vapor pressure of water at 25 $^\circ$C is 3.168 kPa.

\ifthenelse{\equal{\solutions}{true}}{\vspace*{-0.3cm}\input{ch6-solution-9}}{}

\end{enumerate}
