% Problem 8/3 solution
\noindent
\underline{Solution:}\\

\begin{itemize}

\item[a)] This is a reversible process and constant temperature implies that $\Delta\bar{U} = 0$. Also the enthalpy change for an ideal gas depends only on temperature $\Delta\bar{H} = 0$. By using the 1st law and the expression for reversible expansion (see lecture notes), we get:

$$w_{rev} = -RT\ln\left(\frac{V_2}{V_1}\right) = -RT\ln\left(\frac{P_2}{P_1}\right) = - q_{rev}$$

By plugging in the values, we get $w_{rev} = -5.71$ kJ mol$^{-1}$ and $q_{rev} = 5.71$ kJ mol$^{-1}$. Now the definition of entropy gives the entropy change:

$$\Delta\bar{S} = \frac{q_{rev}}{T} = 19.1\textnormal{ J K}^{-1}\textnormal{ mol}^{-1}$$

\item[b)] Irreversible process (free expansion). No external pressure -- no work done ($w = 0$). Thus by the first law $q = 0$ (i.e. no heat exchanged with the surroundings) since $\Delta\bar{U} = 0$ and $\Delta\bar{H} = 0$. Entropy depends only on the endpoints of the path and hence $\Delta\bar{S} = \frac{q_{rev}}{T} = 19.1\textnormal{ J K}^{-1}\textnormal{ mol}^{-1}$, where $q_{rev}$ is from part a). Note that only reversible paths can be used in calculating entropy!

\item[c)] Isolated system, which here means that ``system + surroundings'' is isolated from the rest of the world. For a reversible process we have $dS_{tot} = dS_{syst} + dS_{surr} = 0$. For the system we have:

$$dS_{syst} = \frac{dq_{rev}}{T}\textnormal{ and for the surroundings }dS_{surr} = -\frac{dq_{rev}}{T}$$

Since temperature is constant, $\Delta U = 0$ and $\Delta H = 0$. Therefore:

$$q_{rev} = -w = RT\ln\left(\frac{P_2}{P_1}\right)\textnormal{ (from previous calculations)}$$

Thus $\Delta S_{syst} = R\ln(P_2/P_1$) and $\Delta S_{surr}  = -R\ln(P_2/P_1)$. Thus the total change of entropy (system + surroundings) is zero. Also $q_{tot} = q_{sys} + q_{surr} = RT\ln(P_2/P_1) - RT\ln(P_2/P_1) = 0$. For the same reason, the total work $w_{tot} = w_{sys} + w_{surr} = 0$. Note that the system + surroundings is isolated from the rest of the world and therefore the total change in heat ($q_{tot}$) and work ($w_{tot}$) must clearly be zero.

\item[d)] If we want to calculate the entropy change for the system, we need a reversible path for the calculation. Above this was done: $\Delta S_{syst} = R\ln(P_2/P_1)$. In free expansion, no work is done, $w_{sys} = 0$ (and $w_{surr} = 0$). Since $\Delta U = w_{sys} + q_{sys}$ and $\Delta U = 0$, $q_{sys} = 0$ as well (also then $q_{surr} = 0$). Thus the system is not interacting with the surroundings at all in this process. This means that the entropy of the surroundings does not change either, $\Delta S_{surr} =
0$. The total entropy is then $\Delta S_{tot} = \Delta S_{syst} + \Delta S_{surr} = R\ln(P_2/P_1) = 19.1$ J K$^{-1}$ mol$^{-1}$.

\end{itemize}

\hrule\vspace{0.5cm}
